Unsupervised embeddings capture latent structure of scientific mobility

### 2020-02-25 paper/current/main.tex
edit minor things

% %Organization information
Affiliations mapped to one of 8,661 organizations, disambiguated following a mostly manual process originally designed for the Leiden Rankings of World Universities~\cite{waltman2012leidenrankings}. 
Organizational records were associated with a full name, a type indicating the sector (e.g., University, Government, Industry), and an identifier for the country and city of the organization. 
Sixteen different sector types were included in the analysis, which we aggregated to a four high-level codes: \textit{University}, \textit{Hospital}, \textit{Government}, and \textit{Other}. 
Each record was also associated with a latitude and longitude, however for many organizations these were missing or incorrect. We manually updated the coordinates of 2,267 organizations by searching the institution name and city on Google Maps;
in cases where a precise location of the organization could not be identified, we used the coordinates returned when searching the name of the city.
The data was further enriched with country-level information, including continent, majority language spoken, and language family of the majority language (e.g., the language family of \textit{Spanish} is \textit{Italic}).  
State/province level  information were added using the reverse geocoding service LocationIQ using each organization's latitude and longitude as input. 
Regional census classifications were also added for states in the United States.
For each organization, we also calculated a size as the average number of unique authors (mobile and non-mobile) who published with that organization across each year of our dataset;
in the case that authors publish with multiple affiliations in a single year, they are counted towards each.

As a result of our disambiguation procedure, some affiliations are mapped to two organizations, one specific, and one more general.
For example, any author affiliated with ``Indiana University Bloomington'' will also be listed as being affiliated with the ``Indiana University System'', a more general designation for all public universities in Indiana. 
However, a more general organization may not always occur alongside the more specific one.
For example, a researcher affiliated with the smaller regional school ``Indiana University South Bend'' will be listed as affiliated with only the ``Indiana University System''. 
We identify all specific organizations that always co-occur along a more general one.
For every career trajectory that includes one of these specific organizations, we remove all occurrences of the more general organization;
trajectories containing only a general designation are not altered.

Scientific mobility is complex and can be defined in many possible ways~\cite{robinson2019mobility}.
Here, we consider an author as mobile when at least two distinct organizational affiliations appear in their trajectory during the 2008-2019 time window. 
Under this definition, mobile authors constitute 3,007,192 or 23.2 percent of all authors, and 17,700,095 author-affiliation combinations. 
Mobile authors were associated with, on average, 2.5 distinct organizational affiliations. 
Rates of mobility tended to differ across country. 
For example, France, Qatar, the USA, Iraq, and Luxembourg had the most mobile authors (ref figure).
However, due to their size, the USA, accounted for nearly 40 percent of all mobile authors worldwide, with  10 countries accounting for 80 percent of all mobility (ref figure).
In most cases, countries with a high degree inter-organization mobility also had a high degree of international mobility;
however, the USA and France had a high degree of inter-organization mobility with the corresponding international mobility (ref figure).

@@

%% word2vec
These mobility sentences are used as input to the standard skip-gram negative sampling (SGNS) model for word embedding, commonly known as \textit{word2vec}~\cite{mikolov2013word2vec} . 
In natural-language processing, word embeddings are used to construct dense vector-space representations of words and phrases, in which distance relationships between words correspond to a notion of semantic distance. 
By embedding co-affiliation trajectories, we aim to learn a dense vector for every organization where distance between vectors relates to the tendency for two organizations to co-occur in a token window around the target term in the same career trajectory. 
Vectors are learned by training a neural network to optimize the prediction of so-called context term given a target, in our case using one organization to predict the others appearing in the same co-affiliation trajectory. 
The hidden layer of the trained neural network is used as the

%% Word2vec parameters and training procedure
The SGNS model has several tunable hyper-parameters, including the size of the learned vectors, the size of the token window  considered when training, the minimum frequency threshold for including an organization, and the number of iterations.
We iterate through several values of vector size and window size, repeating analysis for each combination. 
To mitigate the effect of less common organizations, only organizations appearing at least 50 times in mobility sentences are included in the embedding;
this yields 6,580, or 76 percent of all organizations.
We iterate the SGNS training for five epochs over all mobility sentences. 
All word embeddings are trained using the \textit{gensim} python implementation of the SGNS \textit{word2vec} model.

@@

We also calculate two forms of distance between organizations. 
The geographic distance, $g_{ij}$, is the pairwise geographic distance between organizations.
Geographic distance is calculated as the great circle distance, in kilometers, between pairs of organizations. 
When less than one kilometer, a distance of one is imputed. 
The embedding distance, $d_{ij}$, is calculated as the pairwise cosine distance between vectors from the learned organization embedding, where $v_{i}$ and $v_{j}$ represent the vectors and $d_{ij} = 1 - \frac{v_{i} \dot v_{j}}{\lVert v_{i} \rVert \lVert v_{j} \rVert}$.

@@

The gravity law of mobility is known to be sensitive to the case where $f_{ij} = 0$, or zero movement between organizations. 
Here, non-zero flows account for only 4.2 percent of all possible pairs of  the 6,580 organizations represented in the embedding.
This value is comparable to other common applications of the gravity law, such as to hone calls, commuting, and migration \cite{simini2012universal}.
We follow standard practice and exclude from our analysis zero flows.

%
% SUBSECTION: SEMAXIS
%
\subsection*{SemAxis}
By constructing semantic embeddings from mobility trajectories, we allow for unique methodology for understading the embedding space, such as SemAxis.
SemAxis~\cite{an2018semaxis} is a technique that leverages the latent semantic characteristics of word embeddings to represent the position of terms along a conceptual axis, reflecting the relationship of these term to the concept. 
The precise axis is an antonym pair defined \textit{ad-hoc} for the specific use case.
In the case of language, the ends of the axis could be ``good'' and ``bad'', ``surprising'' and ``unsurprising'', or ``masculine'' and ``feminine''. 
In the case of scientific mobility, SemAxis offers a means of operationalizing abstract relationships between organizations.

Here, we apply SemAxis to U.S. universities by defining two axes, one of geography and one of academic prestige. 
The poles of the geographic axis are defined as the mean vector of all vectors corresponding to organizations in California, and then the mean of all vectors of organizations in Massachusetts. 
For the prestige axis, we define a subset of top-ranked universities according to either the Times World University Rankings or based on the mean normalized research impact sourced from the Leiden Rankings. 
The other end of the prestige axis is the geographically-matched (according to census region) set of universities ranked at the bottom of these rankings.
For example, if 20 universities top-ranked universities are selected and six of them are in the Northeastern United States, then the bottom six universities according to the same ranking will be included in the low-ranked subset. 
From the prestige axis we also derive what we call \textit{SemRank}, a university ranking based on the nearness of universities to the \textit{Elite} end of the axis. 
We compare the SemAxis ranking to the traditional university rankings using Spearman rank correlations.

@@

Results are reported for embeddings with a dimensionality of 300 and a window size of one, which were the parameters that produced  that best explained the flux between organizations (see Fig. SI~\ref{fig:supp:hyperparams}). 
We note that this performance and the outcomes of analysis was robust to hyperparemeters settings.

@@

%
% Figure - Gravity law performance
%
\begin{figure}[h!]
	\centering
	\label{fig:gravity_performance}
	\includegraphics[width=\textwidth]{\figdir/GravityLawFig/gravity_performance.pdf}
	\caption{
		\textbf{Embedding distance outperforms and is more robust than geographic distance.}
		\textbf{a.} embedding distance (cosine distance between vectors) explains more of the flux between organization pairs than does geographic distance (\textbf{b}). 
		Red line is line of best fit. 
		Black dots are mean flux across binned distances.
		Color indicates frequency of organization pairs within each hex bin.
		\textbf{c-d.} Predictions  of flux between organization pairs made using embedding distance out-perform those made using geographic distance. 
		Boxplots show distribution of actual flux for binned values of predicted flux.
		Box color corresponds to the degree to which the distribution overlaps $x = y$.
		\textbf{e-f.} Predictions made using model parameters learned from subset of same-country organization pairs. 
		Embedding distance outperforms those made for geographic distance.
		\textbf{g-h.} Embedding distance also outperforms geographic distance when predicting flux using model parameters estimated from only pairs of organizations in different countries. 
		Predictions are not meaningfully different if estimates are learned on all pairs of organizations (Fig. S\ref{fig:supp:gravity_prediction_subsets}).
	}
\end{figure}

We also find that defining the gravity model as a function of embedding distance produces better predictions than does geographic distance.
Comparing predicted against actual flux, we find that using the embedding distance in the gravity function has lower error, and are closer than are predictions made using geographic distance (Fig.~\ref{fig:gravity_performance}c-d). 
Embedding distance also produces better predictions for pairs of organizations within the same country (Fig.~\ref{fig:supp:gravity_correlation}e-f) and for pairs of organizations in different countries (Fig.~\ref{fig:supp:gravity_correlation}g-h). 
We find that these predictions are robust whether we make predictions using a model fit on the subsets themselves (Fig.~\ref{fig:gravity_performance}e-h) or fit on the global data and then applied to the subsets (Fig. SI~\ref{fig:supp:gravity_prediction_subsets}). 
This performance is also much better than that using personalized page rank distance (Fig. SI~\ref{fig:supp:gravity_ppr}).

@@

We observed that word embedding models trained on scientific mobility trajectories offers a reasonable measure of effective distance, one that explains flux between organizations far better than geographic distance. 
Why is this true?
Why does the skip-gram model learn vectors which approximate gravity law relationships?
We find evidence of a latent mathematical connection between the two models, one that makes the skip-gram model applicable to many other forms of mobility trajectory.

@@

In the case of scientific co-affiliation, all movements are without direction (symmetric), meaning that  $f_{ij} = f_{ji}$.
Let's denote $\mathpzc{N}_i$ as the neighborhood set of organization i, which are organizations with non-zero co-affiliation to organization i.
Then, the total flux to or from $i$ can be written as

@@

with the hypothesis that a similar fraction of the scholars are mobile in each organization. Then, our gravity model can be written with only terms from mobility data,

@@

Then, the word embedding model will learns vectors under the condition,

@@

which is restricted to, sum of the conditional probability should be one,

@@

The SGNS model learns an effective distance from trajectories, one which we can visualize the results into two dimensional space by projecting high dimensional organizational vectors into two dimensions vectors using UMAP \cite{mcinnes2018umap} (Fig.~\ref{fig:projection}a).
Overall, organizations are clustered by geography, according to the nations in which they are location;
however, the embedding space also reflects a mix of geographic, historic, cultural, and linguistic relationships between regions at multiple scales.

@@

A unique characteristic of semantic mobility embeddings is the ability to capture structure at multiple scales of granularity.
For example, just as the global embedding captures geographic structure between nations, so too does a projection or organizations in the United States reflect organizational structure based on states and oriented along an East-West axis (Fig.~\ref{fig:projection}d). 
My re-projecting at the state-level for Massachusetts (Fig.~\ref{fig:projection}e), we observe structure between urban Centers (Boston vs. Worcester), but also based on organization type (hospitals tend to be adjacent) and different university systems. Even though UMass Boston is located in Boston, it clusters near other universities in the UMass System, reflecting a relative lack of mobility between the two systems, even within the same state. 
Similar dynamics can be observed in many states (New York, California, etc.), such that prestigious university systems are separate from more regional ones (See SI~?).

\begin{figure}[hp!]
	\centering
	\label{fig:projection}
	\includegraphics[width=\textwidth]{\figdir/Projection/big_embedding_picture.png}
	\caption{
		\textbf{Projection of embedding demonstrates complex multi-scale structure.}
		\textbf{a.}
		UMAP projection \autocite{mcinnes2018umap} of the entire embedding space reveals geographic clustering.
		Each point corresponds to an organization. 
		Size indicates the size of the organization.
		Color indicates the continent. 
		We re-project subsets of organizational vectors to uncover more complex local structure.
		\textbf{b.} Re-projection of organizational vectors roughly corresponding to West-Asian, Southeast Asian, and South Asian countries. 
		\textbf{c.} Re-projection of organizations in Spain, Portugal, and the Portuguese- and Spanish-speaking countries South and Central America. 
		\textbf{d.} Re-projection of organizations in the United States reveal geographic clustering based on state.
		Color indicates Census Bureau-designated regions,
		\textbf{e.} Re-projection of organizations in Massachusetts reveal clustering based on organizational sector, prestige. 
		Color indicates sector designation of organization. 
	}
\end{figure}

@@

We investigate the extent to which geography and language dictate the formation of clusters using \textit{CluSim}~\autocite{gates2019element}.
\textit{CluSim} is a method that calculates the similarity between the cluster results and, for hierarchical clustering, can adjust the relative importance of membership at different levels of the hierarchy with scaling parameter $r$, acting likes ``zooming lens''.  
If $r$ is high, the similarity is based on the lower levels of the dendrogram , and when $r$ is low, the similarity is based on higher levels of the dendrogram. 
In fig. \ref{fig:clustering}a, we report three cluster similarity curves (continent, language, language family) with varying scaling parameter $r$.
As expected, continent plays a major role at higher levels of clustering process (low $r$), language and language family explain the clustering process more at the lower levels (high $r$). 
This suggests that geography, then language, are important factors contributing to the structure of international scientific mobility.

%
% Figure - CLUSTERING
%
\begin{figure}[p!]
	\centering
	\label{fig:clustering}
	\includegraphics[width=\textwidth]{\figdir/Clustering/cluster_country.pdf}
	\caption{
		\textbf{Geography, then language, dictate international mobility.}
		\textbf{a.} Hierarchically clustered similarity matrix of country vectors, aggregated as the mean of all organization vectors within the country that has a more than 25 organizations, excluding the United States. 
		Color of matrix cells corresponds to similarity between country vectors.
		Color of country names corresponds to their cluster.
		Color of three cell rows separated from the matrix correspond to, from left to right, the continent of the country, the language family~\autocite{ethnologue}, and the individual language. 
		\textbf{b.}	\textit{CluSim} \autocite{gates2019element} results demonstrating the factors dictating hierarchical clustering, varying by scaling parameter, $r$. 
		Continent better explains the grouping of country vectors at higher levels of the clustering processes.
		Language family, and then majority spoken language, better explain the grouping of countries at more granular. 
	}
\end{figure}

@@

Using SemAxis, a technique unique to semantic embedding, we reconstruct geographic relationships and prestige hierarchies based defined axis between California and Massachusetts, and between lowly and highly ranked universities according to the Times Ranking of World Universities and the Leiden Ranking field-normalized citation impact (Fig.~\ref{fig:semaxis}). 
We find that, with a few exceptions, all universities in California are nearer the appropriate end of the axis than are Massachusetts universities, and \textit{vice versa}.
Universities in bordering states, such as Arizona and Connecticut, are closer to the ends of the axis (Fig. SI~\ref{fig:supp:semaxis_states}b and Fig. SI~\ref{fig:supp:semaxis_states}e). 
Universities in new York are nearer to the Massachusetts end of the axis (Fig. SI~\ref{fig:supp:semaxis_states}f), whereas universities in Florida are less aligned to either (Fig. SI~\ref{fig:supp:semaxis_states}i).

In terms of prestige, we find prestigious universities towards the top of the axis, such as Stanford, Colombia, MIT, Harvard, and NYU. 
Universities at the bottom of this axis tend to be regional universities with lower national profiles, such as California State University Long Beach, University of West Florida, and Barry University. 
There are strong differences by state in terms of prestige—overall, universities in Massachusetts are nearer the \textit{Elite} end of the axis than are universities in Arizona, Connecticut, Pennsylvania, and Florida (see Fig. SI~\ref{fig:supp:semaxis_states}).

@@

%
% Figure - SemAxis results
%
\begin{figure}[hp!]
	\centering
	\label{fig:semaxis}
	\includegraphics[width=\textwidth]{\figdir/SemAxis/semaxis_fig.pdf}
	\caption{
		\textbf{Embedding captures latent geography and prestige hierarchy.}
		\textbf{a.} illustrates SemAxis projection along two axes, comparing California to Massachusetts universities (left to right), and between the top 20 and geographically-matched bottom 20 universities ranked by the 2018 Times Higher Education ranking of U.S. Universities (bottom to top). 
		Points correspond to organizations, and a sample of individual states are highlighted. 
		\textbf{b.} Comparison between the ranking of organizations by their rank in the Times ranking and the SemAxis rank derived from the similarity to the \textit{Elite} end of the axis prestige axis \textit{a}.
		White points are those top and bottom 20 universities aggregated to define the ends of the axis. 
		The grey box corresponds to the top 20 and bottom 20 ranks. 
	  	The estimate of the spearman rank correlation between the SemAxis and Times ranking is 0.81 (n = 145).
	}
\end{figure}

To what extent are semantic embeddings of mobility actually learning the latent hierarchy of academic mobility?
To test this, we derive a ranking using the prestige axis that we call \textit{SemRank}, and compare them to traditional university rankings. 
SemRank is defined from the position of organizations along the prestige axis, with the top ranked being the furthest towards the \textit{Elite} end of the axis.
We find that SemRank derived is highly correlated with the Times ranking ($\rho = 0.81$, Fig.~\ref{fig:semaxis}b). 
This correlation is robust with the number of organizations used to define the axes (Fig. SI~\ref{fig:supp:semaxis_compare}), such that even using only the single top-ranked and bottom-ranked universities is strongly correlated with the Times Ranking ($\rho = 0.46$). 
For reference, SemRank achieves correlation with the Times Ranking comparable to that between the Times ranking and the Leiden Ranking mean-normalized citation score ($\rho = 0.87$). 
This suggests that word embeddings, in addition to learning the latent geography of mobility, also capture more nuanced characteristics such as latent hierarchy.

>>>>

% %Organization information
Affiliations mapped to one of 8,661 organizations, disambiguated following a mostly manual process originally designed for the Leiden Rankings of World Universities~\cite{waltman2012leidenrankings}. 
Organizational records were associated with a full name, a type indicating the sector (e.g., University, Government, Industry), and an identifier for the country and city of the organization. 
Sixteen different sector types were included in the analysis, which we aggregated to a four high-level codes: \textit{University}, \textit{Hospital}, \textit{Government}, and \textit{Other}. 
Each record was also associated with a latitude and longitude, however, for many organizations, these were missing or incorrect. We manually updated the coordinates of 2,267 organizations by searching the institution name and city on Google Maps;
in cases where a precise location of the organization could not be identified, we used the coordinates returned when searching the name of the city.
The data was further enriched with country-level information, including continent, majority language spoken, and language family of the majority language (e.g., the language family of \textit{Spanish} is \textit{Italic}).  
State/province-level information was added using the reverse geocoding service LocationIQ using each organization's latitude and longitude as input. 
Regional census classifications were also added for states in the United States.
For each organization, we also calculated size as the average number of unique authors (mobile and non-mobile) who published with that organization across each year of our dataset;
in the case that authors publish with multiple affiliations in a single year, they are counted towards each.

As a result of our disambiguation procedure, some affiliations are mapped to two organizations, one specific, and one more general.
For example, any author affiliated with ``Indiana University Bloomington'' will also be listed as being affiliated with the ``Indiana University System'', a more general designation for all public universities in Indiana. 
However, a more general organization may not always occur alongside the more specific one.
For example, a researcher affiliated with the smaller regional school ``Indiana University South Bend'' will be listed as affiliated with only the ``Indiana University System''. 
We identify all specific organizations that always co-occur along with a more general one.
For every career trajectory that includes one of these specific organizations, we remove all occurrences of the more general organization;
trajectories containing only a general designation are not altered.

Scientific mobility is complex and can be defined in many possible ways~\cite{robinson2019mobility}.
Here, we consider an author as mobile when at least two distinct organizational affiliations appear in their trajectory during the 2008-2019 time window. 
Under this definition, mobile authors constitute 3,007,192 or 23.2 percent of all authors and 17,700,095 author-affiliation combinations. 
Mobile authors were associated with, on average, 2.5 distinct organizational affiliations. 
Rates of mobility tended to differ across the country. 
For example, France, Qatar, the USA, Iraq, and Luxembourg had the most mobile authors (ref figure).
However, due to their size, the USA, accounted for nearly 40 percent of all mobile authors worldwide, with  10 countries accounting for 80 percent of all mobility (ref figure).
In most cases, countries with a high degree inter-organization mobility also had a high degree of international mobility;
however, the USA and France had a high degree of inter-organization mobility with the corresponding international mobility (ref figure).

@@

%% word2vec
These mobility sentences are used as input to the standard skip-gram negative sampling (SGNS) model for word embedding, commonly known as \textit{word2vec}~\cite{mikolov2013word2vec} . 
In natural-language processing, word embeddings are used to construct dense vector-space representations of words and phrases, in which distance relationships between words correspond to a notion of semantic distance. 
By embedding co-affiliation trajectories, we aim to learn a dense vector for every organization where the distance between vectors relates to the tendency for two organizations to co-occur in a token window around the target term in the same career trajectory. 
Vectors are learned by training a neural network to optimize the prediction of the so-called context term given a target, in our case using one organization to predict the others appearing in the same co-affiliation trajectory. 
The hidden layer of the trained neural network is used as the

%% Word2vec parameters and training procedure
The SGNS model has several tunable hyper-parameters, including the size of the learned vectors, the size of the token window considered when training, the minimum frequency threshold for including an organization, and the number of iterations.
We iterate through several values of vector size and window size, repeating the analysis for each combination. 
To mitigate the effect of less common organizations, only organizations appearing at least 50 times in mobility sentences are included in the embedding;
this yields 6,580, or 76 percent of all organizations.
We iterate the SGNS training for five epochs overall mobility sentences. 
All word embeddings are trained using the \textit{gensim} python implementation of the SGNS \textit{word2vec} model.

@@

We also calculate two forms of the distance between organizations. 
The geographic distance, $g_{ij}$, is the pairwise geographic distance between organizations.
Geographic distance is calculated as the great circle distance, in kilometers, between pairs of organizations. 
When less than one kilometer, a distance of one is imputed. 
The embedding distance, $d_{ij}$, is calculated as the pairwise cosine distance between vectors from the learned organization embedding, where $v_{i}$ and $v_{j}$ represent the vectors and $d_{ij} = 1 - \frac{v_{i} \dot v_{j}}{\lVert v_{i} \rVert \lVert v_{j} \rVert}$.

@@

The gravity law of mobility is known to be sensitive to the case where $f_{ij} = 0$, or zero movement between organizations. 
Here, non-zero flows account for only 4.2 percent of all possible pairs of the 6,580 organizations represented in the embedding.
This value is comparable to other common applications of the gravity law, such as to hone calls, commuting, and migration \cite{simini2012universal}.
We follow standard practice and exclude from our analysis of zero flows.

%
% SUBSECTION: SEMAXIS
%
\subsection*{SemAxis}
By constructing semantic embeddings from mobility trajectories, we allow for a unique methodology for understanding the embedding space, such as SemAxis.
SemAxis~\cite{an2018semaxis} is a technique that leverages the latent semantic characteristics of word embeddings to represent the position of terms along a conceptual axis, reflecting the relationship of these terms to the concept. 
The precise axis is an antonym pair defined \textit{ad-hoc} for the specific use case.
In the case of language, the ends of the axis could be ``good'' and ``bad'', ``surprising'' and ``unsurprising'', or ``masculine'' and ``feminine''. 
In the case of scientific mobility, SemAxis offers a means of operationalizing abstract relationships between organizations.

Here, we apply SemAxis to U.S. universities by defining two axes, one of geography and one of academic prestige. 
The poles of the geographic axis are defined as the mean vector of all vectors corresponding to organizations in California, and then the mean of all vectors of organizations in Massachusetts. 
For the prestige axis, we define a subset of top-ranked universities according to either the Times World University Rankings or based on the mean normalized research impact sourced from the Leiden Rankings. 
The other end of the prestige axis is the geographically-matched (according to census region) set of universities ranked at the bottom of these rankings.
For example, if 20 universities top-ranked universities are selected and six of them are in the Northeastern United States, then the bottom six universities according to the same ranking will be included in the low-ranked subset. 
From the prestige axis, we also derive what we call \textit{SemRank}, a university ranking based on the nearness of universities to the \textit{Elite} end of the axis. 
We compare the SemAxis ranking to the traditional university rankings using Spearman rank correlations.

@@

Results are reported for embeddings with a dimensionality of 300 and a window size of one, which was the parameters that produced that best explained the flux between organizations (see Fig. SI~\ref{fig:supp:hyperparams}). 
We note that this performance and the outcomes of analysis were robust to hyperparameters settings.

@@

%
% Figure - Gravity law performance
%
\begin{figure}[h!]
	\centering
	\label{fig:gravity_performance}
	\includegraphics[width=\textwidth]{\figdir/GravityLawFig/gravity_performance.pdf}
	\caption{
		\textbf{Embedding distance outperforms and is more robust than geographic distance.}
		\textbf{a.} embedding distance (cosine distance between vectors) explains more of the flux between organization pairs than does geographic distance (\textbf{b}). 
		Red line is the line of best fit. 
		Black dots are mean flux across binned distances.
		The color indicates the frequency of organization pairs within each hex bin.
		\textbf{c-d.} Predictions of flux between organization pairs made using embedding distance out-perform those made using geographic distance. 
		Boxplots show the distribution of actual flux for binned values of predicted flux.
		Box color corresponds to the degree to which the distribution overlaps $x = y$.
		\textbf{e-f.} Predictions made using model parameters learned from the subset of same-country organization pairs. 
		Embedding distance outperforms those made for geographic distance.
		\textbf{g-h.} Embedding distance also outperforms geographic distance when predicting flux using model parameters estimated from only pairs of organizations in different countries. 
		Predictions are not meaningfully different if estimates are learned on all pairs of organizations (Fig. S\ref{fig:supp:gravity_prediction_subsets}).
	}
\end{figure}

We also find that defining the gravity model as a function of embedding distance produces better predictions than does geographic distance.
Comparing predicted against actual flux, we find that using the embedding distance in the gravity function has a lower error, and are closer than are predictions made using geographic distance (Fig.~\ref{fig:gravity_performance}c-d). 
Embedding distance also produces better predictions for pairs of organizations within the same country (Fig.~\ref{fig:supp:gravity_correlation}e-f) and for pairs of organizations in different countries (Fig.~\ref{fig:supp:gravity_correlation}g-h). 
We find that these predictions are robust whether we make predictions using a model fit on the subsets themselves (Fig.~\ref{fig:gravity_performance}e-h) or fit on the global data and then applied to the subsets (Fig. SI~\ref{fig:supp:gravity_prediction_subsets}). 
This performance is also much better than that using personalized page rank distance (Fig. SI~\ref{fig:supp:gravity_ppr}).

@@

We observed that word embedding models trained on scientific mobility trajectories offer a reasonable measure of effective distance, one that explains flux between organizations far better than geographic distance. 
Why is this true?
Why does the skip-gram model learn vectors which approximate gravity law relationships?
We find evidence of a latent mathematical connection between the two models, one that makes the skip-gram model applicable to many other forms of mobility trajectory.

@@

In the case of scientific co-affiliation, all movements are without direction (symmetric), meaning that  $f_{ij} = f_{ji}$.
Let's denote $\mathpzc{N}_i$ as the neighborhood set of organization i, which are organizations with non-zero co-affiliation to the organization i.
Then, the total flux to or from $i$ can be written as

@@

with the hypothesis that a similar fraction of the scholars is mobile in each organization. Then, our gravity model can be written with only terms from mobility data,

@@

Then, the word embedding model will learn vectors under the condition,

@@

which is restricted to, a sum of the conditional probability should be one,

@@

The SGNS model learns an effective distance from trajectories, one which we can visualize the results into two-dimensional space by projecting high dimensional organizational vectors into two dimensions vectors using UMAP \cite{mcinnes2018umap} (Fig.~\ref{fig:projection}a).
Overall, organizations are clustered by geography, according to the nations in which they are located;
however, the embedding space also reflects a mix of geographic, historic, cultural, and linguistic relationships between regions at multiple scales.

@@

A unique characteristic of semantic mobility embeddings is the ability to capture structure at multiple scales of granularity.
For example, just as the global embedding captures geographic structure between nations, so too does a projection or organization in the United States reflect organizational structure based on states and oriented along an East-West axis (Fig.~\ref{fig:projection}d). 
My re-projecting at the state-level for Massachusetts (Fig.~\ref{fig:projection}e), we observe structure between urban Centers (Boston vs. Worcester) but also based on organization type (hospitals tend to be adjacent) and different university systems. Even though UMass Boston is located in Boston, it clusters near other universities in the UMass System, reflecting a relative lack of mobility between the two systems, even within the same state. 
Similar dynamics can be observed in many states (New York, California, etc.), such that prestigious university systems are separate from more regional ones (See SI~?).

\begin{figure}[hp!]
	\centering
	\label{fig:projection}
	\includegraphics[width=\textwidth]{\figdir/Projection/big_embedding_picture.png}
	\caption{
		\textbf{Projection of embedding demonstrates complex multi-scale structure.}
		\textbf{a.}
		UMAP projection \autocite{mcinnes2018umap} of the entire embedding space reveals geographic clustering.
		Each point corresponds to an organization. 
		Size indicates the size of the organization.
		The color indicates the continent. 
		We re-project subsets of organizational vectors to uncover more complex local structures.
		\textbf{b.} Re-projection of organizational vectors roughly corresponding to West-Asian, Southeast Asian, and South Asian countries. 
		\textbf{c.} Re-projection of organizations in Spain, Portugal, and the Portuguese- and Spanish-speaking countries South and Central America. 
		\textbf{d.} Re-projection of organizations in the United States reveal geographic clustering based on state.
		The color indicates Census Bureau-designated regions,
		\textbf{e.} Re-projection of organizations in Massachusetts reveal clustering based on the organizational sector, prestige. 
		Color indicates sector designation of organization. 
	}
\end{figure}

@@

We investigate the extent to which geography and language dictate the formation of clusters using \textit{CluSim}~\autocite{gates2019element}.
\textit{CluSim} is a method that calculates the similarity between the cluster results and, for hierarchical clustering, can adjust the relative importance of membership at different levels of the hierarchy with scaling parameter $r$, acting likes ``zooming lens''.  
If $r$ is high, the similarity is based on the lower levels of the dendrogram, and when $r$ is low, the similarity is based on higher levels of the dendrogram. 
In fig. \ref{fig:clustering}a, we report three cluster similarity curves (continent, language, language family) with varying scaling parameter $r$.
As expected, continent plays a major role at higher levels of clustering process (low $r$), language and language family explain the clustering process more at the lower levels (high $r$). 
This suggests that geography, then language, are important factors contributing to the structure of international scientific mobility.

%
% Figure - CLUSTERING
%
\begin{figure}[p!]
	\centering
	\label{fig:clustering}
	\includegraphics[width=\textwidth]{\figdir/Clustering/cluster_country.pdf}
	\caption{
		\textbf{Geography, then language, dictate international mobility.}
		\textbf{a.} Hierarchically clustered similarity matrix of country vectors, aggregated as the mean of all organization vectors within the country that has more than 25 organizations, excluding the United States. 
		The color of matrix cells corresponds to the similarity between country vectors.
		Color of country names corresponds to their cluster.
		Color of three cell rows separated from the matrix correspond to, from left to right, the continent of the country, the language family~\autocite{ethnologue}, and the individual language. 
		\textbf{b.}	\textit{CluSim} \autocite{gates2019element} results demonstrating the factors dictating hierarchical clustering, varying by scaling parameter, $r$. 
		Continent better explains the grouping of country vectors at higher levels of the clustering processes.
		Language family, and then majority spoken language, better explain the grouping of countries at more granular. 
	}
\end{figure}

@@

Using SemAxis, a technique unique to semantic embedding, we reconstruct geographic relationships and prestige hierarchies based defined axis between California and Massachusetts, and between lowly and highly ranked universities according to the Times Ranking of World Universities and the Leiden Ranking field-normalized citation impact (Fig.~\ref{fig:semaxis}). 
We find that, with a few exceptions, all universities in California are nearer the appropriate end of the axis than are Massachusetts universities, and \textit{vice versa}.
Universities in bordering states, such as Arizona and Connecticut, are closer to the ends of the axis (Fig. SI~\ref{fig:supp:semaxis_states}b and Fig. SI~\ref{fig:supp:semaxis_states}e). 
Universities in New York are nearer to the Massachusetts end of the axis (Fig. SI~\ref{fig:supp:semaxis_states}f), whereas universities in Florida are less aligned to either (Fig. SI~\ref{fig:supp:semaxis_states}i).

In terms of prestige, we find prestigious universities towards the top of the axis, such as Stanford, Colombia, MIT, Harvard, and NYU. 
Universities at the bottom of this axis tend to be regional universities with lower national profiles, such as California State University Long Beach, University of West Florida, and Barry University. 
There are strong differences by the state in terms of prestige—overall, universities in Massachusetts are nearer the \textit{Elite} end of the axis than are universities in Arizona, Connecticut, Pennsylvania, and Florida (see Fig. SI~\ref{fig:supp:semaxis_states}).

@@

%
% Figure - SemAxis results
%
\begin{figure}[hp!]
	\centering
	\label{fig:semaxis}
	\includegraphics[width=\textwidth]{\figdir/SemAxis/semaxis_fig.pdf}
	\caption{
		\textbf{Embedding captures latent geography and prestige hierarchy.}
		\textbf{a.} illustrates SemAxis projection along two axes, comparing California to Massachusetts universities (left to right), and between the top 20 and the geographically-matched bottom 20 universities ranked by the 2018 Times Higher Education ranking of U.S. Universities (bottom to top). 
		Points correspond to organizations, and a sample of individual states are highlighted. 
		\textbf{b.} Comparison between the ranking of organizations by their rank in the Times ranking and the SemAxis rank derived from the similarity to the \textit{Elite} end of the axis prestige axis \textit{a}.
		White points are those top and bottom 20 universities aggregated to define the ends of the axis. 
		The gray box corresponds to the top 20 and bottom 20 ranks. 
	  	The estimate of the spearman rank correlation between the SemAxis and Times ranking is 0.81 (n = 145).
	}
\end{figure}

To what extent are semantic embeddings of mobility actually learning the latent hierarchy of academic mobility?
To test this, we derive a ranking using the prestige axis that we call \textit{SemRank}, and compare them to traditional university rankings. 
SemRank is defined from the position of organizations along the prestige axis, with the top-ranked being the furthest towards the \textit{Elite} end of the axis.
We find that SemRank derived is highly correlated with the Times ranking ($\rho = 0.81$, Fig.~\ref{fig:semaxis}b). 
This correlation is robust with the number of organizations used to define the axes (Fig. SI~\ref{fig:supp:semaxis_compare}), such that even using only the single top-ranked and bottom-ranked universities is strongly correlated with the Times Ranking ($\rho = 0.46$). 
For reference, SemRank achieves correlation with the Times Ranking comparable to that between the Times ranking and the Leiden Ranking mean-normalized citation score ($\rho = 0.87$). 
This suggests that word embeddings, in addition to learning the latent geography of mobility, also capture more nuanced characteristics such as latent hierarchy.
